Validation of marker gene expression

### 2025-02-26 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
account for cell line samples

---
title: "Validation of marker gene expression"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000001"
---

@@

```{r}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_dt <- consensus_dt |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  merge(validation_groups_dt, by = "consensus_annotation", all.x = TRUE) |> 
  merge(gene_exp_dt, by = c("barcodes", "library_id"), all.x = TRUE) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  merge(markers_dt, by = "ensembl_gene_id", all.x = TRUE, allow.cartesian = TRUE)

@@

```{r}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_dt |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  unique() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

```{r}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_dt |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_dt, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent expressed
    percent_exp = (detected_count/total_cells) * 100,
    all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100, 
    # account for NA/unknowns and set axes order
    broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |> 
      factor(levels =  c(unique(markers_dt$validation_group_annotation), "unknown"))
  )

@@

```{r}
# get list of celltypes to keep and assign colors 
celltype_groups <- group_stats_df |> 
  dplyr::pull(broad_celltype_group) |> 
  unique() |>
  as.character()

@@

```{r}
# filter markers to those that are actually relevant 
# we will only plot the marker genes for cell types that are part of the assigned broad validation group for this project
# we don't care about plotting marker genes for cell types that aren't present here 
# note that we will use this for both the dotplot and the heatmap
filtered_markers_df <- markers_dt |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% group_stats_df$gene_symbol)

@@

```{r, fig.height=10, fig.width=15}
# filter out low expressed genes
dotplot_df <- unique_gene_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
  dplyr::arrange(broad_celltype_group) |> 
  # add a label for the plot 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
  # remove marker genes that aren't present in final annotations and set x axis order 
  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
  dplyr::mutate(
    # set orders of gene symbol and validation groups 
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
  )

@@

```{r}
# filter low expressed genes 
grouped_dotplot_df <- group_stats_df |> 
  dplyr::select(broad_celltype_group, validation_group_annotation, all_markers_mean_exp, all_markers_percent_exp, total_cells) |>
  unique() |> 
  dplyr::filter(all_markers_mean_exp > 0, all_markers_percent_exp > 5, validation_group_annotation %in% celltype_groups) |>
  dplyr::arrange(broad_celltype_group) |> 
  # make sure labels are correct 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})")),
                validation_group_annotation = factor(validation_group_annotation, levels = celltype_order))

>>>>

---
title: "Validation of marker gene expression"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  project_id: "SCPCP000020"
---

@@

```{r, results='asis'}
# check for if the project is all cell lines or not 
is_cell_line <- all(consensus_dt$sample_type == "cell line")

if(is_cell_line){
  glue::glue("
    <div class=\"alert alert-info\">

    All samples in this project are from cell lines. 
    No cell typing was performed, so no plots will be shown.

    </div>
  ")
}
```

```{r, eval=!is_cell_line}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_dt <- consensus_dt |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  merge(validation_groups_dt, by = "consensus_annotation", all.x = TRUE) |> 
  merge(gene_exp_dt, by = c("barcodes", "library_id"), all.x = TRUE) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  merge(markers_dt, by = "ensembl_gene_id", all.x = TRUE, allow.cartesian = TRUE)

@@

```{r, eval=!is_cell_line}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_dt |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  unique() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

```{r, eval=!is_cell_line}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_dt |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_dt, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent expressed
    percent_exp = (detected_count/total_cells) * 100,
    all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100, 
    # account for NA/unknowns and set axes order
    broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |> 
      factor(levels =  c(unique(markers_dt$validation_group_annotation), "unknown"))
  )

@@

```{r, eval=!is_cell_line}
# get list of celltypes to keep and assign colors 
celltype_groups <- group_stats_df |> 
  dplyr::pull(broad_celltype_group) |> 
  unique() |>
  as.character()

@@

```{r, eval=!is_cell_line}
# filter markers to those that are actually relevant 
# we will only plot the marker genes for cell types that are part of the assigned broad validation group for this project
# we don't care about plotting marker genes for cell types that aren't present here 
# note that we will use this for both the dotplot and the heatmap
filtered_markers_df <- markers_dt |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% group_stats_df$gene_symbol)

@@

```{r, eval=!is_cell_line, fig.height=10, fig.width=15}
# filter out low expressed genes
dotplot_df <- unique_gene_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
  dplyr::arrange(broad_celltype_group) |> 
  # add a label for the plot 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
  # remove marker genes that aren't present in final annotations and set x axis order 
  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
  dplyr::mutate(
    # set orders of gene symbol and validation groups 
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
  )

@@

```{r, eval=!is_cell_line}
# filter low expressed genes 
grouped_dotplot_df <- group_stats_df |> 
  dplyr::select(broad_celltype_group, validation_group_annotation, all_markers_mean_exp, all_markers_percent_exp, total_cells) |>
  unique() |> 
  dplyr::filter(all_markers_mean_exp > 0, all_markers_percent_exp > 5, validation_group_annotation %in% celltype_groups) |>
  dplyr::arrange(broad_celltype_group) |> 
  # make sure labels are correct 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})")),
                validation_group_annotation = factor(validation_group_annotation, levels = celltype_order))
